---
title: "Tooth Growth Analysis"
author: "Mark Culp"
date: "March 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview:
This analysis examines the effects of vitamin C on 
tooth growth in guinea pigs.  Each of the 60 guinea 
pigs studied received one of three doses of vitamin
C per day: 0.5, 1.0, or 2.0 mg/day.  There were two 
delivery methods: orange juice (OJ) and ascorbic 
acid (VC).  The response measures the growth in 
odontoblasts (cells responsible for tooth growth).


```{r}

# Load dataset
library(datasets)
data("ToothGrowth")

```

## Exploratory analysis

```{r}

# List length of rows / columns
dim(ToothGrowth)

# Determine column data types
str(ToothGrowth)

# Range of tooth growth
range(ToothGrowth$len)

# Levels and counts of delivery methods/supplements
table(ToothGrowth$supp)

# Levels and counts of dosages
table(ToothGrowth$dose)


```

## Data summary 

## Comparison of tooth growth by supp and dose

## Conclusions and assumptions








