---
title: "Tooth Growth Analysis"
author: "Mark Culp"
date: "March 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview:
This analysis examines the effects of vitamin C on 
tooth growth in guinea pigs.  Each of the 60 guinea 
pigs studied received one of three doses of vitamin
C per day: 0.5, 1.0, or 2.0 mg/day.  There were two 
delivery methods: orange juice (OJ) and ascorbic 
acid (VC).  The response measures the growth in 
odontoblasts (cells responsible for tooth growth).


```{r}

# Load dataset
library(datasets)
data("ToothGrowth")

```

## Exploratory analysis

```{r}

# List length of rows / columns
dim(ToothGrowth)

# Determine column data types
str(ToothGrowth)

# Range of tooth growth
range(ToothGrowth$len)

# Levels and counts of delivery methods/supplements
table(ToothGrowth$supp)

# Levels and counts of dosages
table(ToothGrowth$dose)


```

## Data summary 

```{r}

# Summarize columns 
summary(ToothGrowth)


# Let's look at our 30 guinea pigs given orange
# juice and ascorbic acid
oj <- ToothGrowth[ToothGrowth$supp == "OJ",]
vc <- ToothGrowth[ToothGrowth$supp == "VC",]

# A histogram of tooth odontoblast length
# appears to have a positive slope possibly 
# due to increasing dosages of orange juice. 
hist(oj$len)

# The histogram of the tooth growth for guinea
# pigs given ascorbic acid looks almost normal.
hist(vc$len)

```

## Comparison of tooth growth by supp and dose

```{r}
# So lets compare tooth growth by supplement by 
# dosage levels

# Given an 0.5 milligram/day dosage, we have
# 10 samples of guinea pigs given orange juice
# and 10 samples of guinea pigs given ascorbic acid
oj05 <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "OJ",]
vc05 <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "VC",]
dim(oj05)
dim(vc05)

# Our means and standard error for the orange
# juice samples at the 0.5 mg/day level are:
mean(oj05$len)
sd(oj05$len)

# So we conduct a t-test on the orange 
# juice sample at the 0.5 mg/day dosage level.
# This shows we can reject the null hypothesis
# that the orange juice had no impact on the 
# guinea pigs' tooth growth.
t.test(oj05$len)

# Our means and standard error for the ascorbic
# acid samples at the 0.5 mg/day level are:
mean(vc05$len)
sd(vc05$len)

# So we conduct a t-test on the ascorbic 
# acid sample at the 0.5 mg/day dosage level.
# This shows we can reject the null hypothesis
# that the ascorbic acid had no impact on the 
# guinea pigs' tooth growth.
t.test(vc05$len)

# --------------------------------

# We now compare the guinea pigs' tooth growth 
# under both supplements: 
# ----------------------
# At the 0.5 mg/daydosage level
tg05 <- ToothGrowth[ToothGrowth$dose == 0.5,] 
t.test(len ~ supp, paired = F, var.equal = F, data = tg05)

# At the 1.0 mg/daydosage level
tg10 <- ToothGrowth[ToothGrowth$dose == 1.0,] 
t.test(len ~ supp, paired = F, var.equal = F, data = tg10)

# At the 2.0 mg/daydosage level
tg20 <- ToothGrowth[ToothGrowth$dose == 2.0,] 
t.test(len ~ supp, paired = F, var.equal = F, data = tg20)




# The power of the orange juice samples at the
# 0.5 mg/day level are:
# power.t.test(n=10,delta = 13.23,sd= 4.459709, type="one.sample", alt = "one.sided")$power

```

## Conclusions and assumptions

```{r}


```





